Icicle: write once, run once

### 2016-05-24 2016fhpc/Main.tex
Work on intro
Reorg to move counter example to the intro

\authorinfo{
  Amos Robinson$^\dagger$$^\ddagger$
  \and Ben Lippmeier$^\dagger$
}{
  \vspace{5pt}
  \shortstack{
    $^\dagger$Computer Science and Engineering \\
    University of New South Wales, Australia \\[2pt]
    \textsf{amosr,benl@cse.unsw.edu.au}
  }
  \shortstack{
    $^\ddagger$Ambiata      \\
    Grow with data          \\[2pt]
    \textsf{amos.robinson@ambiata.com}
  }
}

@@

\begin{abstract}
We present a pure query language, Icicle, which statically guarantees that multiple queries over the same input data set will be fused. We include a modal type system that guarantees these fused results will be evaluated with a single pass over the input data. Icicle queries are compiled via a fold-based intermediate language to efficient C code. We present production benchmarks demonstrating significant speedup over existing queries written in R, as well as the widely used Unix tools \texttt{grep} and \texttt{wc}.

@@

\category
	{D.3.4}
	{Programming Languages}
	{Processors---Compilers; Optimization}

\terms
	Languages, Performance

\keywords
	Arrays; Fusion

>>>>

\authorinfo{
  Amos Robinson$^{\alpha \beta}$
  \and Ben Lippmeier$^{\beta \gamma}$
}{
  \vspace{5pt}
  \shortstack{
    $^\alpha$Ambiata      \\
    Grow with data          \\[2pt]
    \textsf{amos.robinson@ambiata.com}
  }
  \shortstack{
    $^\beta$UNSW Australia\\
    \\[2pt]
    ~~~~\textsf{amosr,benl@cse.unsw.edu.au}
  }
  \shortstack{
    $^\gamma$Vertigo Technology \\
    \\[2pt]
    ~~\textsf{benl@vergo.co}
  }
}

@@

\begin{abstract}
We present Icicle, a pure streaming query language which statically guarantees that multiple queries over the same input data stream will be fused. Icicle queries are compiled via a fold-based intermediate language to efficient C code. We present production benchmarks demonstrating significant speedup over existing queries written in R, as well as the widely used Unix tools \texttt{grep} and \texttt{wc}.

@@

### 2016-05-24 2016fhpc/section/01-Introduction.tex
Reorg to move counter example to the intro

% \ben{Start with the example.}
Suppose we have a table, @stocks@, containing daily open and close prices for a set of companies. We wish to find the number of days where the open price exceeds the close price, and vice versa:
\begin{code}
  SELECT COUNT(*) FROM stocks WHERE open > close;
  SELECT COUNT(*) FROM stocks WHERE open < close;
\end{code}
As both queries are over the same input data set, we would like to evaluate them using a single pass over this data. As they use different @WHERE@ clause filters, we must fuse them by lifting the filter predicate into the expression being evaluated:

\begin{code}
  SELECT COUNT(IF(open > close, 1, 0))
       , COUNT(IF(open < close, 1, 0))
  FROM stocks;
\end{code}
Suppose we also want to compute the mean of open price on those days:
\begin{code}
  SELECT SUM(open) / COUNT(*)
  FROM   stocks
  WHERE  open > close;
\end{code}
Now we can perform a similar trick to fuse all three queries together.
%
\begin{code}
  SELECT COUNT(IF(open > close, 1, 0))
       , COUNT(IF(open < close, 1, 0))
       , SUM  (IF(open > close, open, 0))
       / COUNT(IF(open > close, 1, 0))
FROM stocks;
\end{code}

Fusing SQL queries by hand is tedious and error-prone. At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple separate queries over hundreds of gigabytes worth of data. For such applications we \emph{must} automatically fuse these queries and be sure that the result is correct in terms of both its value and performance charateristics.

In the fused query above, note that we are counting the number of times where the open price exceeds the close price twice: once for the first source query and once for the third. In practice, the result of query fusion tends to have many common sub expressions, and we must also guarantee that the duplicates will be eliminated.

\ben{Mention resumable queries in intro. The contributions mention them.}

\ben{Rewrite the examples using this syntax?}
In Icicle, the above queries are expressed as:
\begin{code}
table stocks { open : Int; close : Int; }
query more = filter open > close in count;
query less = filter open < close in count;
query mean = filter open > close in sum open / count;
\end{code}

These queries are then fused together, common subexpression elimination is performed, and efficient C code is generated.

The contributions of this paper are:
\begin{itemize}
\item
we present a type system that guarantees any two queries on the same input table can be fused, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};

@@

>>>>

At Ambiata, we peform feature generation for machine learning applications by executing many thousands of simple queries over hundreds of gigabytes worth of data.\footnote{In 2016 this was a lot of data.} For such applications we must automatically fuse these separate queries and be sure that the result can be executed in a single pass over the input. We also ingest multiple gigabytes of new data per day, and must incrementally update existing features without recomputing them all from scratch. Our feature generation process is performed in parallel on hundreds of nodes of a cloud based system, and if we performed neither fusion or incremental update then the cost of the computation would begin to exceed the salaries of the developers.

For example queries, suppose we have a table, @stocks@, containing daily open and close prices for a set of companies. We want to compute the number of days where the open price exceeded the close price, and vice versa. We also want the mean of the open price for days in which the open price exceeded the close price. In Icicle we write the three sub queries as follows:

\begin{code}
table stocks { open : Int, close : Int }
query 
  more  = filter open > close in count;
  less  = filter open < close in count;
  mean  = filter open > close in sum open / count;
\end{code}
In the above code, @open > close@ and @close < open@ are the filter predicates, and @count@ counts the number of times the body of each filter construct was entered. We describe this construct more fully in \REF

A joint query like this can be converted to a back-end language such a SQL, but doing it by hand it tedious and error prone. As the three sub queries use different filter predicates we cannot use a single @SELECT@ statement and a @WHERE@ expression to implement the filter. We instead need to lift each predicate to an expression-level conditional:
%
\begin{code}
  SELECT COUNT(IF(open > close, 1, 0))
       , COUNT(IF(open < close, 1, 0))
       , SUM  (IF(open > close, open, 0))
       / COUNT(IF(open > close, 1, 0))
  FROM stocks;
\end{code}

As we see, the result of query fusion also tends to have many common sub expressions, and we wish to guarantee that the duplicates in the fused result are eliminated.

Joint queries such as the stocks example can be evaluated in a streaming, incremental fashion. As a counter example, suppose we have a table with two fields @key@ and @value@, and we wish to find the mean of values whose key matches the last one in the table. We might try something like:
%
\begin{code}
  table kvs { key : Int; value : Int }
  query avg = let k = last key
              in  filter (key == k) in mean value;
\end{code}

Unfortunately, although the \emph{result} we desire is computable, the \emph{algorithm} for doing so implied by the above query cannot be evaluated in a streaming fashion. When we are streaming through the table we always have access to the last key in the stream, but to perform the filter to select rows that match this key would require that we stream through the table again from the start. We need a better solution. With this in mind, the contributions of this paper are:
\begin{itemize}
\item
we present a type system that guarantees any two queries on the same input table can be fused, as well as allowing queries to be resumed as new data is received\sref{s:IcicleSource};

@@

% \ben{Rewrite the examples using this syntax?}
% In Icicle, the above queries are expressed as:
% \begin{code}
% table stocks { open : Int; close : Int; }
% query more = filter open > close in count;
% query less = filter open < close in count;
% query mean = filter open > close in sum open / count;
% \end{code}

% \ben{Start with the example.}
% \begin{code}
%   SELECT COUNT(*) FROM stocks WHERE open > close;
%   SELECT COUNT(*) FROM stocks WHERE open < close;
% \end{code}
% As both queries are over the same input data set, we would like to evaluate them using a single pass over this data. As they use different @WHERE@ clause filters, we must fuse them by lifting the filter predicate into the expression being evaluated:

% \begin{code}
%   SELECT COUNT(IF(open > close, 1, 0))
%        , COUNT(IF(open < close, 1, 0))
%   FROM stocks;
% \end{code}
% Suppose we also want to compute the mean of open price on those days:
% \begin{code}
%   SELECT SUM(open) / COUNT(*)
%   FROM   stocks
%   WHERE  open > close;
% \end{code}

% These queries are then fused together, common subexpression elimination is performed, and efficient C code is generated.

### 2016-05-24 2016fhpc/section/02-Source.tex
typing rules
Work on intro
Reorg to move counter example to the intro

%!TEX root = ../Main.tex
\section{Icicle}
\label{s:IcicleSource}

The table definition in $\mi{Table}$ gives a table name and the names and types of columns.
$\mi{Exp}$ defines the expressions used as query and function bodies.
$\mi{Prim}$ defines the primitive types.
Function definitions are given in $\mi{Fun}$.
Query definitions are given in $\mi{Query}$.
$\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.

$\mi{Fun}$ is used for function definitions.
The @count@ function takes no arguments and returns a fold counting the number of elements.
\begin{code}
function count
 = fold c = 0 then c + 1;
\end{code}

$\mi{Query}$ defines an aggregate query which must return a single value, and gives it a name:
\begin{code}
query count_all_records = count;
\end{code}

@@

All Icicle queries must execute in a single pass over the data, as reading the data multiple times is expensive. 
Ensuring that queries only require a single pass over the data also allows us to resume queries from where they left off, when new data arrives.

@@

Here is an example to highlight one benefit of tracking these modalities.
Suppose we have a table with two fields: key and value.
We wish to find the mean of values whose key matches the most recent key.
Someone new to stream programming might write the following query, forgetting that they can only read the data once: first find the most recent key, then go back and filter all the data to only those with the right key.

\begin{code}
   let k    = newest key
in let avg  = filter (key == k) in mean value
in avg
\end{code}

@@

>>>>

%!TEX root = ../Main.tex

\eject
\section{Elements and Aggregates}

All Icicle queries must execute in a single pass over the data. Ensuring that queries only require a single pass over the data also allows us to resume queries from where they left off, when new data arrives. However, not all queries can be evaluated in a single pass. For example, suppose we have a table with two fields key and value, and we wish to find the mean of values whose key matches the last one in the table. We might try something like:

\begin{code}
  table t { key : Int; value : Int }
  query avg = let k    = last key
              in  filter (key == k) in mean value;
\end{code}
The problem with the above query is that it requires two passes through the data. One to

Someone new to stream programming might write the following query, forgetting that they can only read the data once: first find the most recent key, then go back and filter all the data to only those with the right key.

@@

@@

Here is an example to highlight one benefit of tracking these modalities.

@@

\section{Source Language}
\label{s:IcicleSource}

The table definition in $\mi{Table}$ gives a table name and the names and types of columns.
$\mi{Exp}$ defines the expressions used as query and function bodies.
$\mi{Prim}$ defines the primitive types.
Function and query definitions are given in $\mi{Def}$.
$\mi{Top}$ is the top-level program, which specifies a table, the set of function bindings, and the set of queries.

$\mi{Def}$ is used for function and query definitions.
The @count@ function takes no arguments and returns a fold counting the number of elements.
\begin{code}
function count
 = fold c = 0 then c + 1;
\end{code}

Queries must return a single value, and are given it a name:
\begin{code}
query count_all_records = count;
\end{code}
